Revisiting Short-Chain Fatty Acids and the Microbiota in Colorectal Cancer

### 2018-05-23 submission/manuscript.Rmd
add intro template

@@

Our observations conssitently found no difference in SCFA concentration between individuals with normal colons and patients with either adenoma or carcinoma tumors. Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence. Overall, these results suggest that the SCFAs produced by resident microbes do not protect against tumor. By focusing on other types of metabolites, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related cause of death within the US []. The envrionment has been attributed as an important risk factor for CRC []. This has lead to the investigation of how the microbiota may exacerbate or cause tumorigensis [] and whether the bacterial community is altered []. Multiple reports in case/control studies on the microbiota and CRC have identified taxa commonly associated with the normal colon to be decreased in patients with carcinoma tumors []. Many of these taxa actively produce short-chain fatty acids (SCFAs) from fiber that is part of our general diet []. The most extensively studied of these SCFAs are acetate, butyrate, and propionate []. Overall, CRC is a growing health problem and the microbiota through SCFAs have shown promise as a viable treatment option.

Within model systems SCFAs, in particular butyrate, have shown positive results in CRC treatment []. In particular, butyrate has been shown to inhibit cancer cell growth in *in vitro* systems []. Additionally, supplementation with food sources that bacteria use to create these SCFAs may also be able to confer beneficial effects. Fiber supplementation in mouse models of CRC caused an overall reduction in tumor burden while increasing SCFA concentrations []. Although these model systems provide important preliminary evidence towards the ability of SCFAs to reduce and treat tumors, the studies reporting benefit in humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA concentrations to protect against CRC in human populations. The initial case/control studies that investigated SCFA concentrations in CRC have found that patients with carcinomas had lower concentrations of acetate, butyrate, and propionate versus patients with adenomas and individuals with normal colons []. Although this would argue that increasing SCFA concentrations could be protective against tumorigenesis, fiber supplementation in randomized controlled trials have consistently failed to protect against tumor recurrence []. These findings would argue against the utility of treatments that aim to use SCFAs to reduce or protect against tumorigenesis. Given the lack of clear evidence in human studies of the benefit of SCFAs in CRC there is a need for more investigation into this area.

Our current study will fill some of the current gaps in the literature that relate to the study of SCFAs and CRC in human populations. Specifically, it will attempt to replicate previous case/control findings on SCFA concentrations in individuals with and without tumors. Additionally, by increasing the adenoma tumor cohort it we will be able to identify if the observation of a continuous gradual reduction in SCFA concentrations occurs as tumor severity increases. It also will build upon these observations and assess the utility of using SCFAs and Operational Taxonomic Units (OTUs) as a potential risk stratification tool of colorectal tumors. This study will provide important information on the replicability of previous findings in humans by extensively studying how SCFAs could be associated with colorectal tumors.

To accomplish this task we will directly measure the concentration of acetate, butyrate, and propionate within fecal samples for two different groups. The first group will be at a cross sectional point in time while the second group will be before (pre-) and after (post-)treatment. Additionally, we also will assess whether adding SCFA concentrations to OTU data improves the classification of adenoma or carcinoma, use PICRUSt and metagenomic sequencing to assess the genes involved in SCFA synthesis, and analyze how well 16S rRNA gene sequencing can predict SCFA concentrations. This extensive investigation will provide additional information as to whether the taxa that are decreased in carcionma use SCFAs to help protect against tumorigenesis.

@@

Our observations found no difference in SCFA concentration, their utility as a classification tool, and for genes for enzymes involved in SCFA synthesis between individuals with normal colons and patients with either adenoma or carcinoma tumors. Although these results are different than other reports in the literature, they do align with the randomized controlled trials that have tested fiber use in preventing colorectal tumor recurrence. Overall, these results suggest that the SCFAs produced by resident microbes do not protect against tumor. By focusing on other types of metabolites, the identification of more promising therapeutic options for use in treating colorectal cancer may be found.

### 2018-05-23 submission/manuscript.tex
add intro template

>>>>

Colorectal cancer (CRC) is currently the third leading cancer-related
cause of death within the US {[}{]}. The envrionment has been attributed
as an important risk factor for CRC {[}{]}. This has lead to the
investigation of how the microbiota may exacerbate or cause tumorigensis
{[}{]} and whether the bacterial community is altered {[}{]}. Multiple
reports in case/control studies on the microbiota and CRC have
identified taxa commonly associated with the normal colon to be
decreased in patients with carcinoma tumors {[}{]}. Many of these taxa
actively produce short-chain fatty acids (SCFAs) from fiber that is part
of our general diet {[}{]}. The most extensively studied of these SCFAs
are acetate, butyrate, and propionate {[}{]}. Overall, CRC is a growing
health problem and the microbiota through SCFAs have shown promise as a
viable treatment option.

Within model systems SCFAs, in particular butyrate, have shown positive
results in CRC treatment {[}{]}. In particular, butyrate has been shown
to inhibit cancer cell growth in \emph{in vitro} systems {[}{]}.
Additionally, supplementation with food sources that bacteria use to
create these SCFAs may also be able to confer beneficial effects. Fiber
supplementation in mouse models of CRC caused an overall reduction in
tumor burden while increasing SCFA concentrations {[}{]}. Although these
model systems provide important preliminary evidence towards the ability
of SCFAs to reduce and treat tumors, the studies reporting benefit in
humans has been less convincing.

There is a lack of evidence on the benefit of increasing SCFA
concentrations to protect against CRC in human populations. The initial
case/control studies that investigated SCFA concentrations in CRC have
found that patients with carcinomas had lower concentrations of acetate,
butyrate, and propionate versus patients with adenomas and individuals
with normal colons {[}{]}. Although this would argue that increasing
SCFA concentrations could be protective against tumorigenesis, fiber
supplementation in randomized controlled trials have consistently failed
to protect against tumor recurrence {[}{]}. These findings would argue
against the utility of treatments that aim to use SCFAs to reduce or
protect against tumorigenesis. Given the lack of clear evidence in human
studies of the benefit of SCFAs in CRC there is a need for more
investigation into this area.

Our current study will fill some of the current gaps in the literature
that relate to the study of SCFAs and CRC in human populations.
Specifically, it will attempt to replicate previous case/control
findings on SCFA concentrations in individuals with and without tumors.
Additionally, by increasing the adenoma tumor cohort it we will be able
to identify if the observation of a continuous gradual reduction in SCFA
concentrations occurs as tumor severity increases. It also will build
upon these observations and assess the utility of using SCFAs and
Operational Taxonomic Units (OTUs) as a potential risk stratification
tool of colorectal tumors. This study will provide important information
on the replicability of previous findings in humans by extensively
studying how SCFAs could be associated with colorectal tumors.

To accomplish this task we will directly measure the concentration of
acetate, butyrate, and propionate within fecal samples for two different
groups. The first group will be at a cross sectional point in time while
the second group will be before (pre-) and after (post-)treatment.
Additionally, we also will assess whether adding SCFA concentrations to
OTU data improves the classification of adenoma or carcinoma, use
PICRUSt and metagenomic sequencing to assess the genes involved in SCFA
synthesis, and analyze how well 16S rRNA gene sequencing can predict
SCFA concentrations. This extensive investigation will provide
additional information as to whether the taxa that are decreased in
carcionma use SCFAs to help protect against tumorigenesis.
